### 2016-05-11 Projects/DelCo_2014_2015/Reports/Hose_Stream_Tests/TeX_Files/DelCo_Hose_Stream_Tests.tex
Minor edits to hose stream and gas burner reports

Fire suppression tactics using hose streams also affect the ventilation in a structure and may impact the movement of smoke and heat through a structure.  As vents are made to advance the hose line or if ventilation inducing handline tactics are in practice, hoselane impacts on venmtilation may be even more significant. Additional research addressing the coordination of suppression tactics and the impact on ventilation is needed to complete recommendations on fire control tactics to appropriate standards, education, and training documents.

There exists a multitude of different tools, methods, and techniques used to apply water for fire suppression using hose streams. A common type of nozzle that is used to apply water during fire suppression in residential structures is an adjustable-pattern spray nozzle, commonly referred to as a combination nozzle~\cite{NFPA_1964}. Most combination nozzles contain an adjustable tip that can be rotated to produce different stream patterns. The types of streams produced by a combination nozzle are often divided into three categories based on the angle of the stream with respect to the nozzle: a straight stream typically produces an angle less than 15\SI{}{\degree}, a narrow fog stream produces an angle between 15\SI{}{\degree} and 45\SI{}{\degree}, and a wide fog stream produces an angle between 45\SI{}{\degree} to 80\SI{}{\degree}~\cite{IFSTA:Essentials_of_FF} \textit{(cited 5th edition because that's where I got this info from, I'm assuming I can change it to 6th?)}. Increasing the angle of the stream allows for a larger area to be covered by the water spray.  However, increasing the angle decreases the distance reached by the stream and its penetration power. 
%\textit{Essentials of Fire Fighting} defines 
%https://docs.google.com/presentation/d/1hZ5DZhYXxCkcIEHNZKumH48b_Q53iM2D7asXkMYlVMM/embed?hl=en&size=l&slide=id.p39

%http://www.tft.com/literature/library/files/bfa.pdf
%Keith Royer, "Water for Fire Fighting", (Engineering Extension Service, Iowa State University, Ames, Iowa) p. 6 
In addition to the type of stream pattern used, the method used to apply the hose stream can also vary. For example, Royer suggested using a clockwise rotation instead of a counterclockwise application pattern for the following reasons~\cite{Royer:ISU}:
\begin{enumerate} 
	\item Clockwise rotation drives most of the heated gases, smoke, and flames away from the nozzle. Counterclockwise rotation does just the opposite.
	\item Steam formed by the clockwise rotation has a violent rolling action. Counterclockwise rotation produces steam with an inactive and lazy action.
	\item A clockwise roation increases the efficiency of water and produces a faster knockdown time than counterclockwise rotation.
\end{enumerate}

\section{Objectives}
\label{sec:objectives}
This report is concerned with examining the impact of hose stream and application pattern, or nozzle motion, selection on air movement inside a full scale test sturctures built to represent residential sized structures. Numerous experiments containing different flow path configurations were conducted to study the impact of hose stream and water application selection on ventilation in a structure. More specifically, the experiments were conducted to fulfill the following objectives:
\begin{enumerate}
	\item Determine which hose stream patterns and application patterns cause significant air movement in a structure 
	\item Determine the amount of air movement induced by different hose streams and application patterns

\end{enumerate}
Four types of hose streams and four different water application patterns were studied throughout the experiments described in this report. Three types of hose stream patterns using a combination nozzle were studied: straight stream, narrow fog stream, and wide fog stream patterns. Water was flowed into the structure by having the hose in a static, or fixed, position, by moving the hose line left-to-right across the room in a sweeping motion, and by rotating the hose line in both the clockwise and counterclockwise directions. Additionally, a set of experiments comparing the impact of a straight stream from a combination nozzle and a solid stream from a smooth bore nozzle on air flows in a structure was conducted. Images of the four types of hose streams are presented in Fig.~\ref{fig:hose_streams}.

\begin{figure}[!ht]
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/NF_example}
		\caption{Narrow fog stream}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/WF_example}
		\caption{Wide fog stream}
	\end{subfigure}
	\\~\\
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/SS_70}
		\caption{Straight stream from combination nozzle}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/SB_70}
		\caption{Solid stream from smooth bore nozzle}
	\end{subfigure}
	\caption[Hose stream patterns used during experiments.]{The four types of hose stream patterns used during water flow experiments.}
	\label{fig:hose_streams}
\end{figure}

@@

% =========================
% = EXPERIMENTAL OVERVIEW =
% =========================
\chapter{Experimental Overview}
\label{chap:exp_overview}
To study the impact of different hose stream and application pattern selections on air movement inside a structure, various water flow experiments were conducted in two experimental structures designed to replicate typical residential structures. The experiments focused on studying the differences in amount of air movement caused by specific combinations of four different hose stream patterns and four different water application patterns.

\section{Experimental Setup}
\label{sec:exp_setup}
The series of field experiments described in this report were conducted in two structures of similar design located at the Delaware County Emergency Services Training Center (ESTC) in Sharon Hill, PA. Differential pressure and temperature sensors were installed at ventilation points throughout the residential-sized test structures to determine air velocity during the water flow experiments.

@@

\begin{figure}[!ht]
	\includegraphics[width=5.25in]{../Figures/Pictures/east_structure}
	\\~\\
	\includegraphics[width=5.25in]{../Figures/Pictures/west_structure}
	\caption[North side of the Single Story and Two Story Structures.]{North side of the Single Story (top) and Two Story (bottom) Test Structures.}
	\label{fig:struct_pics}
\end{figure}

The interior walls of the first floor of each structure were framed with steel studs set to 0.40~m (16~in) centers and track and were lined with 13~mm (0.5~in) thick cement board. The walls were composed of 16~mm (0.625~in) Type X gypsum board. Additionally, the ceiling was composed of two layers of 13~mm (0.5~in) thick cement board.
\FloatBarrier

The first floor ceiling support of each structure was composed of wood truss joist I-beams (TJIs) with a 299~mm (11.75~in) depth. Each TJI was composed of laminated veneer lumber flanges with a cross section of 29~mm (1.125~in) x 44~mm (1.75~in) and an 11~mm (0.43~in) thick oriented strand board web as shown in Fig.~\ref{fig:TJI}. Tongue and grove, 18.3~mm (0.72~in) thick, oriented strand board was screwed to the top of the TJIs.

@@

The second floor of the Two Story Structure was built on the wood ceiling support described above and was connected to the first floor by a stairwell. The second story walls were wood framed with nominal 51~mm (2~in) by 102~mm (4~in) studs set to 0.40~m (16~in) centers. The interior walls were protected by 16~mm (0.63~in) fire rated gypsum board, 16~mm (0.63~in) durarock board, and a second layer of 16~mm (0.625~in) fire rated gypsum board. The exterior walls were protected with 11~mm (0.44~in) oriented strand board and 8~mm (0.31~in) fiber cement lap siding.

\textit{Should the english units be in fractions instead of decimals? For example, 0.625 --> 5/8. And if decimals, should they all be rounded to same place?}
\textit{check dimensions in Fig 2.4, are they consistant with text.  Upstairs is slightly larger and downstairs?}

\subsubsection{Layout}
\label{sec:layout}
Dimensioned floor plans of the Single Story and Two Story Structures are presented in Figures~\ref{fig:east_dimensioned_plan}~and~\ref{fig:west_dimensioned_plan}, respectively.

@@

The interior dimensions of each structure were approximately 6.1~m (20~ft) wide, 11~m (36~ft) long and 2.4~m (8~ft) high. The stairs connecting the two floors of the Two Story Structure started 1.6~m (5.25~ft) off the south wall with a width of 1.2~m (4~ft) off the east wall.

The exterior doorways of each structure and the stairwell doorway on the second level of the Two Story Structure all contained doors that were opened or closed at certain instances to change the ventilation patterns of the structure. All other doorways in the structures did not contain a door, so to close the doorway, a sheet of gypsum board was used to cover the opening and the doorway remained closed for the duration of the test procedure.
\FloatBarrier

\subsection{Instrumentation}
\label{sec:instrumentation}
 For every experiment, gas velocity at an interior doorway was calculated using differential pressure transducers connected to bi-directional velocity probes~\cite{McCaffrey:Combustion_and_Flame}. An array of six bi-directional probes located at the northeast interior doorway of the Single Story Structure is pictured below in Fig.~\ref{fig:BDPs}. A single thermocouple was attached to each bi-directional probes to measure the gas temperature to be used in conjunction with the pressure measurement to derive velocity. Every thermocouple attached to the bi-directional probes was exposed-bead, Chromel-Alumel (type K) with a 1.0~mm (0.04 in) diameter. Starting with the exposed bead, the thermocouple wire was sheathed in a 3.2~mm (0.13 in) diameter Inconel shield that was 0.76~m (2.5~ft) in length.

The approximate locations of the bi-directional probes are annotated in the plan views of the Single Story Structure and Two Story Structure experimental setups (Fig.~\ref{fig:east_setup}--\ref{fig:flow_path_2}) found in Section~\ref{sec:exp_procedure}. To minimize potential measurement errors due to external environmental conditions (i.e., wind), only velocity data measured by bi-directional probes located at interior doorways are considered in this report. The set of bi-directional probes at the northeast interior doorway of the Single Story Structure, denoted as ``A6'', contained six probes located at distances of 0.30~m (1.0~ft), 0.60~m (2.0~ft), 0.90~m (3.0~ft), 1.20~m (3.9~ft), 1.50~m (4.9~ft), and 1.80~m (5.9~ft) below the soffit of the doorway. The set of bi-directional probes at the interior doorway in the Two Story Structure, denoted as ``A10'', contained eight probes located at distances of 0.08~m (0.25~ft), 0.34~m (1.1~ft), 0.61~m (2.0~ft), 0.88~m (2.9~ft), 1.15~m (3.7~ft), 1.42~m (4.7~ft), 1.68~m (5.5~ft), and 1.95~m (6.4~ft) below the soffit of the corresponding doorway.

@@

Bi-directional probes with pressure transducers and single thermocouples were used to measure the gas velocity. Bare-bead Type K thermocouples were co-located with the probe. A gas velocity measurement study, examining the doorway flow of pre-flashover compartment fires, yielded expanded uncertainty measurements ranging from $\pm0.14$ to $\pm0.22$ for bi-directional probes of similar design~\cite{Bryant:FSJ2009}. The total expanded uncertainty for gas velocity in these experiments is estimated to be $\pm18$\%.

Water flow rate was measured with a pressure and flow meter combination shown in Fig.~\ref{fig:flow_meter}. The meter consists of a section of 6.35~cm (2.5~in) cast aluminum pipe with a 0--4.1 MPa (0--600 psi) pressure transducer and a paddlewheel type flow sensor with a range of 0 to 4800~lpm (1250~gpm). The pressure transducer and paddlewheel both connect to the battery operated control box where the pressure transducer voltage is converted to a pressure and the paddlewheel pulse count is converted to a volumetric flow rate. The manufacturer reports a $\pm5$\% calibration expanded uncertainty for the flow sensor and $\pm3$\%  for the pressure sensor~\cite{Akron}. The pressure transducer was calibrated with a known analog pressure gauge. The flow meter was calibrated by capturing water over time and measuring that mass of water to determine the flow rate. The total expanded uncertainty was estimated at $\pm10$\%. 
% FIND FLOW METER REFERENCE
\begin{figure}[!ht]
	\includegraphics[width=4in]{../Figures/Pictures/flow_meter}
	\caption[Flow meter used to measure flow rate during experiments.]{Pressure and flow meter combination used to measure water flow rate during the experiments.}
	\label{fig:flow_meter}
\end{figure}
\FloatBarrier

\section{Experimental Procedure}
\label{sec:exp_procedure}
Every experiment discussed in this report involved discharging water at a set flow rate from a combination nozzle in a straight, narrow fog, or wide fog stream pattern or from a smooth bore nozzle in a solid stream pattern. Specific doors were opened or closed before and/or during the experiments to change the ventilation in the structure.

A total of seven scenarios were conducted to study the impact of different hose stream and application patterns on gas movement in a structure; three used a combination nozzle attached to a monitor to flow water at approximately 120~GPM, three used a combination nozzle attached directly to a hoseline to flow water at approximately 120~GPM (Fig.~\ref{fig:monitor+handline}), and one used combination and smooth bore nozzles attached to a monitor to flow water at approximately 180~GPM. Experiments that used a monitor to flow water always applied the stream in a fixed position and thus were focused on changing the location of water application in addition to the stream pattern, while the experiments that used a combination nozzle attached directly to a manual hoseline  to flow water focused on varying the pattern in which water was applied in addition to the stream pattern.

\textit{Should I present flow rates in GPM only, or LPM (GPM)? What about the tables? As of now, GPM is used, but that goes against the rule of keeping the type of units consistent in a table. Same goes for ``1.75 in'' handline.}
\begin{figure}[!ht]
	\minipage{3in}
	\begin{center}
		\includegraphics[width=2.9in]{../Figures/Pictures/monitor}
	\end{center} 
	\endminipage \hfill
	\minipage{3in}
	\begin{center}
		\includegraphics[width=2.9in]{../Figures/Pictures/handline}
	\end{center}
	\endminipage \hfill
	\caption[Monitor and handline equipped with a combination nozzle.]{Monitor (left) and 1.75~in handline (right) equipped with a combination nozzle and used to flow water during experiments.}
	\label{fig:monitor+handline}
\end{figure}
\FloatBarrier

Three different structure layouts were used for the seven tests. The tests conducted in the Single Story Structure, Tests 33 and 34, were performed using an identical structure layout, presented in Fig.~\ref{fig:east_setup}. Test 33 used a monitor and combination nozzle to flow water, and Test 34 used a combination nozzle attached directly to a handline to flow water.

The main difference between the two layouts used for the Two Story Structure experiments is the position of the south side door on the first level. Tests 16 and 18 used the configuration with the door opened (Fig.~\ref{fig:flow_path_1}), and Tests 17, 19, and 70 used the configuration with the door closed (Fig.~\ref{fig:flow_path_2}). Tests 16 and 17 used a combination nozzle attached to a monitor to flow water; Tests 18 and 19 used a combination nozzle attached directly to a handline to flow water; and Test 70 used a combination nozzle and a smooth bore nozzle attached to a monitor to flow water in a straight and solid stream, respectively.

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/Specific_Tests/East_Hose_Test_Annotated}
	\caption[Plan view of the Single Story Structure setup for Tests 33 and 34.]{Plan view of the Single Story Structure setup for Tests 33 and 34. The view is annotated with the approximate location of water flow (nozzle graphic), the direction of the established flow path (red line), and the approximate location of bi-directional probe array A6 (blue square).}
	\label{fig:east_setup}
\end{figure}
\FloatBarrier

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/Specific_Tests/West_Hose_Test_2nd_Floor_Annotated}
	\\~\\
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/Specific_Tests/West_Hose_Test_18_1st_Floor_Annotated}
	\caption[Plan view of the Two Story Structure setup for Tests 16 and 18.]{Plan view of the second floor (top) and first floor (bottom) setup in the Two Story Structure for Tests 16 and 18. The approximate location of water flow (nozzle graphic), the direction of the established flow path (red line), and the approximate location of bi-directional probe array A10 (blue square). The stairwell door and north side, west double door on the second floor were opened and closed at certain instances during the tests; the flow path was considered fully established when both first and second floor doors were in the opened position.}
	\label{fig:flow_path_1}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/Specific_Tests/West_Hose_Test_2nd_Floor_Annotated}
	\\~\\
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/Specific_Tests/West_Hose_Test_19_1st_Floor_Annotated}
	\caption[Plan view of the Two Story Structure setup for Tests 17 and 19.]{Plan view of the second floor (top) and first floor (bottom) setup in the Two Story Structure for Tests 17 and 19. The approximate location of water flow (nozzle graphic), the direction of the established flow path (red line), and the approximate location of bi-directional probe array A10 (blue square). The stairwell door and north side, west double door on the second floor were opened and closed at certain instances during the tests; the flow path was considered fully established when both first and second floor doors were in the opened position.}
	\label{fig:flow_path_2}
\end{figure}
\FloatBarrier

\subsection{Monitor Experiments}
\label{sec:monitor_procedure}
Four different scenarios were conducted using a monitor equipped with a combination nozzle to flow water at 120~GPM in a straight stream, narrow fog stream, and wide fog stream pattern. One test, Test 33, occurred in the Single Story Structure, and the two other tests, Tests 16 and 17, occurred in the Two Story Structure. A fourth test, Test 70, was conducted in the Two Story Structure and used a combination nozzle and a smooth bore nozzle attached to a monitor to flow water at 180~GPM in a straight stream and solid stream, respectively. The monitors were set to flow water from a fixed position.  Water was always applied in a fixed position during the monitor experiments. Thus, the monitor experiments primarily focused on the impact of different hose stream patterns on air movement in a structure and provided a baseline to compare the impact of the motion of the nozzle on the air movement through the stucture.

>>>>

Fire suppression tactics using hose streams also affect the ventilation in a structure and may impact the movement of smoke and heat through a structure. As vents are made to advance the hoseline or if ventilation inducing handline tactics are in practice, hose stream impacts on ventilation may be even more significant. Additional research addressing the coordination of suppression tactics and the impact on ventilation is needed to complete recommendations on fire control tactics to appropriate standards, education, and training documents.

There exists a multitude of different tools, methods, and techniques used to apply water for fire suppression using hose streams. A common type of nozzle that is used to apply water during fire suppression in residential structures is an adjustable-pattern spray nozzle, commonly referred to as a combination nozzle~\cite{NFPA_1964}. Most combination nozzles contain an adjustable tip that can be rotated to produce different stream patterns. The types of streams produced by a combination nozzle are often divided into three categories based on the angle of the stream with respect to the nozzle: a straight stream typically produces an angle less than 15\SI{}{\degree}; a narrow fog stream produces an angle between 15\SI{}{\degree} and 45\SI{}{\degree}; and a wide fog stream produces an angle between 45\SI{}{\degree} and 80\SI{}{\degree}~\cite{IFSTA:Essentials_of_FF}. Increasing the angle of the stream allows for a larger area to be covered by the water spray.  However, increasing the angle decreases the distance reached by the stream and its penetration power. 
%\textit{Essentials of Fire Fighting} defines 
%https://docs.google.com/presentation/d/1hZ5DZhYXxCkcIEHNZKumH48b_Q53iM2D7asXkMYlVMM/embed?hl=en&size=l&slide=id.p39

%http://www.tft.com/literature/library/files/bfa.pdf
%Keith Royer, "Water for Fire Fighting", (Engineering Extension Service, Iowa State University, Ames, Iowa) p. 6 
In addition to the type of stream pattern used, the method used to apply the hose stream can also vary. For example, Royer suggested using a clockwise rotation instead of a counterclockwise application pattern for the following reasons~\cite{Royer:ISU}:
\begin{enumerate} 
	\item Clockwise rotation drives most of the heated gases, smoke, and flames away from the nozzle, while counterclockwise rotation does just the opposite.
	\item Steam formed by the clockwise rotation has a violent rolling action. Counterclockwise rotation produces steam with an inactive and lazy action.
	\item A clockwise rotation increases the efficiency of water and produces a faster knockdown time than counterclockwise rotation.
\end{enumerate}

\section{Objectives}
\label{sec:objectives}
This report is concerned with examining the impact of hose stream and application pattern, or nozzle motion, selection on air movement inside two full scale test structures built to represent residential sized structures. Numerous experiments containing different flow path configurations were conducted to study the impact of hose stream and water application selection on ventilation in a structure. More specifically, the experiments were conducted to fulfill the following objectives:
\begin{enumerate}
	\item Determine which hose stream patterns and application patterns cause significant air movement in a structure 
	\item Determine the amount of air movement induced by different hose streams and application patterns

\end{enumerate}
Four types of hose streams and four different water application patterns were studied throughout the experiments described in this report. Three types of hose stream patterns using a combination nozzle were studied: straight stream, narrow fog stream, and wide fog stream. Water was flowed into the structure by having the hose in a static, or fixed, position; by moving the hoseline left-to-right across the room in a sweeping motion; and by rotating the hoseline in both the clockwise and counterclockwise directions. Additionally, a set of experiments comparing the impact of a straight stream from a combination nozzle and a solid stream from a smooth bore nozzle equipped with a 1~in tip on air flows in a structure was conducted. Images of the four types of hose streams are presented in Fig.~\ref{fig:hose_streams}.

\begin{figure}[!ht]
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/NF_example}
		\caption{Narrow fog stream from combination nozzle}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/WF_example}
		\caption{Wide fog stream from combination nozzle}
	\end{subfigure}
	\\~\\
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/SS_70}
		\caption{Straight stream from combination nozzle}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/SB_70}
		\caption{Solid stream from smooth bore nozzle with 1~tip}
	\end{subfigure}
	\caption[Hose stream patterns used during experiments.]{The four types of hose stream patterns used during water flow experiments.}
	\label{fig:hose_streams}
\end{figure}

@@

% =========================
% = EXPERIMENTAL OVERVIEW =
% =========================
\chapter{Experimental Overview}
\label{chap:exp_overview}
To study the impact of different hose stream and application pattern selections on air movement inside a structure, various water flow experiments were conducted in two experimental structures designed to replicate typical residential structures. The experiments focused on studying the differences in the amount of air movement caused by specific combinations of four different hose stream patterns and four different water application patterns.

\section{Experimental Setup}
\label{sec:exp_setup}
The series of field experiments described in this report were conducted in two structures of similar design located at the Delaware County Emergency Services Training Center (ESTC) in Sharon Hill, PA. Differential pressure and temperature sensors were installed at ventilation points throughout the residential sized test structures to determine air velocity during the water flow experiments.

@@

\begin{figure}[!ht]
	\includegraphics[width=5.25in]{../Figures/Pictures/east_structure}
	\\~\\
	\includegraphics[width=5.25in]{../Figures/Pictures/west_structure}
	\caption[North side of the Single Story and Two Story Structures.]{North side of the Single Story (top) and Two Story (bottom) Structures.}
	\label{fig:struct_pics}
\end{figure}

The interior walls of the first floor of each structure were framed with steel studs set to 400~mm (16~in) centers and track and were lined with 13~mm (0.5~in) thick cement board. The walls were composed of 16~mm (0.63~in) Type X gypsum board. Additionally, the ceiling was composed of two layers of 13~mm (0.5~in) thick cement board.
\FloatBarrier

The first floor ceiling support of each structure was composed of wood truss joist I-beams (TJIs) with a 298~mm (11.75~in) depth. Each TJI was composed of laminated veneer lumber flanges with a cross section of 29~mm (1.13~in) x 44~mm (1.75~in) and an 11~mm (0.43~in) thick oriented strand board web as shown in Fig.~\ref{fig:TJI}. Tongue and grove, 18.3~mm (0.72~in) thick, oriented strand board was screwed to the top of the TJIs.

@@

The second floor of the Two Story Structure was built on the wood ceiling support described above and was connected to the first floor by a stairwell. The walls of the second floor were wood framed with nominal 51~mm (2~in) by 102~mm (4~in) studs set to 400~mm (16~in) centers. The interior walls were protected by 16~mm (0.63~in) fire rated gypsum board, 16~mm (0.63~in) durarock board, and a second layer of 16~mm (0.63~in) fire rated gypsum board. The exterior walls were protected with 11~mm (0.44~in) oriented strand board and 8~mm (0.31~in) fiber cement lap siding.

\subsubsection{Layout}
\label{sec:layout}
Dimensioned floor plans of the Single Story and Two Story Structures are presented in Fig.~\ref{fig:east_dimensioned_plan} and Fig.~\ref{fig:west_dimensioned_plan}, respectively.

The ceiling height of each structure was approximately 2.4~m (8~ft). The interior dimensions of the Single Story Structure were approximately 6.1~m (20~ft) by 11~m (36~ft), and the interior dimensions of the first and second floors of the Two Story Structure were 5.8~m (19~ft) by 10.7~m (35.1~ft) and 6.1~m (20~ft) by 10.9~m (35.8~ft), respectively. The stairs connecting the two floors of the Two Story Structure started 1.6~m (5.3~ft) off the south wall with a width of 1.2~m (4~ft) off the east wall.

The exterior doorways of each structure and the stairwell doorway on the second level of the Two Story Structure all contained doors that were opened or closed at certain instances during tests to change the ventilation pattern of the structure. All other doorways in the structures did not contain a door. So, in order to close these doorways, a sheet of gypsum board was used to cover the opening, and the doorway remained closed for the duration of the test procedure.
\clearpage

@@

\subsection{Instrumentation}
\label{sec:instrumentation}
For every experiment, gas velocity at an interior doorway was calculated using differential pressure transducers connected to bi-directional velocity probes~\cite{McCaffrey:Combustion_and_Flame}. An array of six bi-directional probes located at the northeast interior doorway of the Single Story Structure is pictured below in Fig.~\ref{fig:BDPs}. A single thermocouple was attached to each bi-directional probes to measure the gas temperature to be used in conjunction with the pressure measurement to derive velocity. Every thermocouple attached to the bi-directional probes was an exposed-bead, Chromel-Alumel (type K) with a 1.0~mm (0.04 in) diameter. The thermocouple wire was sheathed in a 3.2~mm (0.13 in) diameter Inconel shield that started at the exposed bead and was 0.76~m (2.5~ft) in length.

The approximate locations of the bi-directional probes are annotated in the plan views of the experimental setups for the Single Story and Two Story Structures (Figs.~\ref{fig:east_setup}--\ref{fig:flow_path_2}) in Section~\ref{sec:exp_procedure}. To minimize potential measurement errors due to external environmental conditions (i.e., wind), only velocity data measured by bi-directional probes located at interior doorways are considered in this report. The set of bi-directional probes at the northeast interior doorway of the Single Story Structure, denoted as ``A6'' and pictured in Fig.~\ref{fig:BDPs}, contained six probes located at distances of 0.30~m (1.0~ft), 0.60~m (2.0~ft), 0.90~m (3.0~ft), 1.20~m (3.9~ft), 1.50~m (4.9~ft), and 1.80~m (5.9~ft) below the soffit of the doorway. The set of bi-directional probes at the interior doorway in the Two Story Structure, denoted as ``A10'', contained eight probes located at distances of 0.08~m (0.25~ft), 0.34~m (1.1~ft), 0.61~m (2.0~ft), 0.88~m (2.9~ft), 1.15~m (3.7~ft), 1.42~m (4.7~ft), 1.68~m (5.5~ft), and 1.95~m (6.4~ft) below the soffit of the corresponding doorway.

@@

Bi-directional probes with pressure transducers and single thermocouples were used to measure the gas velocity. Bare-bead, type K thermocouples were co-located with the probe. A gas velocity measurement study examining the doorway flow of pre-flashover compartment fires yielded expanded uncertainty measurements ranging from $\pm0.14$ to $\pm0.22$ for bi-directional probes of similar design~\cite{Bryant:FSJ2009}. The total expanded uncertainty for gas velocity in these experiments is estimated to be $\pm18$\%.

Water flow rate was measured with a pressure and flow meter combination shown in Fig.~\ref{fig:flow_meter}. The meter consists of a section of 63.5~mm (2.5~in) cast aluminum pipe with a 0--4.1 MPa (0--600 psi) pressure transducer and a paddlewheel type flow sensor with a range of 0 to 4800~lpm (1250~gpm). The pressure transducer and paddlewheel both connect to the battery operated control box where the pressure transducer voltage is converted to a pressure and the paddlewheel pulse count is converted to a volumetric flow rate. The manufacturer reports a $\pm5$\% calibration expanded uncertainty for the flow sensor and $\pm3$\% for the pressure sensor~\cite{Akron:2009}. The pressure transducer was calibrated with a known analog pressure gauge. The flow meter was calibrated by capturing water over time and measuring that mass of water to determine the flow rate. The total expanded uncertainty was estimated at $\pm10$\%. 
% FIND FLOW METER REFERENCE
\begin{figure}[!ht]
	\includegraphics[width=4in]{../Figures/Pictures/flow_meter}
	\caption[Flow meter used to measure flow rate during experiments.]{Pressure and flow meter combination used to measure water flow rate during the experiments.}
	\label{fig:flow_meter}
\end{figure}
\FloatBarrier

\section{Experimental Procedure}
\label{sec:exp_procedure}
Every experiment discussed in this report involved discharging water at a set flow rate from a combination nozzle in a straight, narrow fog, or wide fog stream pattern or from a smooth bore nozzle with a 1~in tip in a solid stream pattern. Specific doors were opened or closed before and/or during the tests to change the ventilation within the structure.

A total of seven experiments were conducted to study the impact of different hose stream and application patterns on gas movement in a structure; three used a combination nozzle attached to a monitor to flow water at approximately 120~gpm, three used a combination nozzle attached directly to a 1.75~in handline to flow water at approximately 120~gpm (Fig.~\ref{fig:monitor+handline}), and one used both a combination nozzle and a smooth bore nozzle attached to a monitor to flow water at approximately 180~gpm. Experiments that used a monitor to flow water always applied the stream in a fixed position and were focused on changing the location of water application in addition to the stream pattern, while the experiments that used a combination nozzle attached directly to a 1.75~in handline to flow water focused on varying the pattern in which water was applied in addition to the stream pattern.

\begin{figure}[!ht]
	\minipage{3in}
	\begin{center}
		\includegraphics[width=2.9in]{../Figures/Pictures/monitor}
	\end{center} 
	\endminipage \hfill
	\minipage{3in}
	\begin{center}
		\includegraphics[width=2.9in]{../Figures/Pictures/handline}
	\end{center}
	\endminipage \hfill
	\caption[Monitor and handline equipped with a combination nozzle.]{Monitor (left) and 1.75~in handline (right) equipped with a combination nozzle and used to flow water during experiments.}
	\label{fig:monitor+handline}
\end{figure}
\FloatBarrier

Three different structure layouts were used for the seven experiments. Tests 1 and 5 were conducted in the Single Story Structure and used the layout shown in Fig.~\ref{fig:east_setup}. Test 1 used a monitor and combination nozzle to flow water, and Test 5 used a combination nozzle attached directly to a handline to flow water.

Two different layouts were used for experiments conducted in the Two Story Structure. The main difference between the two layouts was the position of the south side door on the first level. Tests 2 and 6 used a configuration with the door opened (Fig.~\ref{fig:flow_path_1}), while Tests 3, 4, and 7 used a configuration with the door closed (Fig.~\ref{fig:flow_path_2}). Tests 2 and 3 used a combination nozzle attached to a monitor to flow water; Tests 6 and 7 used a combination nozzle attached directly to a handline to flow water; and Test 4 used both a combination nozzle and a 1~in tip smooth bore nozzle attached to a monitor to flow water in a straight and solid stream, respectively.

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/Specific_Tests/East_Hose_Test_Annotated}
	\caption[Plan view of the Single Story Structure setup for Tests 1 and 5.]{Plan view of the Single Story Structure setup for Tests 1 and 5. The view is annotated with the approximate location of water flow (nozzle graphic), the direction of the established flow path (red line), and the approximate location of bi-directional probe array A6 (blue square). The north side double doors remained opened for the entire duration of the experiments.}
	\label{fig:east_setup}
\end{figure}
\FloatBarrier

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/Specific_Tests/West_Hose_Test_2nd_Floor_Annotated}
	\\~\\
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/Specific_Tests/West_Hose_Test_18_1st_Floor_Annotated}
	\caption[Plan view of the Two Story Structure setup for Tests 2 and 6.]{Plan view of the second floor (top) and first floor (bottom) setup in the Two Story Structure for Tests 2 and 6. The view is annotated with the approximate location of water flow (nozzle graphic), the direction of the established flow path (red line), and the approximate location of bi-directional probe array A10 (blue square). The stairwell door and the west double door on the north side of the second floor were opened and closed at certain instances during the tests.}
	\label{fig:flow_path_1}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/Specific_Tests/West_Hose_Test_2nd_Floor_Annotated}
	\\~\\
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/Specific_Tests/West_Hose_Test_19_1st_Floor_Annotated}
	\caption[Plan view of the Two Story Structure setup for Tests 3 and 7.]{Plan view of the second floor (top) and first floor (bottom) setup in the Two Story Structure for Tests 3, 4, and 7. The view is annotated with the approximate location of water flow (nozzle graphic), the direction of the established flow path (red line), and the approximate location of bi-directional probe array A10 (blue square). The stairwell door and the west double door on the north side of the second floor were opened and closed at certain instances during the tests.}
	\label{fig:flow_path_2}
\end{figure}
\FloatBarrier

\subsection{Monitor Experiments}
\label{sec:monitor_procedure}
Three experiments, Tests~1--3, used a monitor equipped with a combination nozzle to flow water at 120~gpm in a straight stream, narrow fog stream, and wide fog stream. Test 1 occurred in the Single Story Structure, and Tests 2 and 3 occurred in the Two Story Structure. A fourth experiment, Test 4, was conducted in the Two Story Structure and used a combination nozzle and a smooth bore nozzle attached to a monitor to flow water at 180~gpm in a straight stream and solid stream, respectively. The monitor was always set to flow water from a fixed position. Thus, the monitor experiments primarily focused on the impact of different hose stream patterns on air movement in a structure and provided a baseline to compare the impact of the motion of the nozzle on the air movement through the structure.

Test 1 utilized a monitor equipped with a combination nozzle to flow water from Room C of the Single Story Structure to the area above the doorway on the south wall of Room B (Fig.~\ref{fig:test_1_pic}). A plan view of the structure's layout annotated with the approximate location of the bi-directional probe array used to measure gas velocity through an interior doorway and the general direction of the fully established flow path for Test 1 can be found in Fig.~\ref{fig:east_setup}. The procedure for Test 1 was the simplest of all the test procedures. First, water was flowed in a straight stream pattern for 30 seconds, then the stream was changed to a narrow fog for 30 seconds, and finally the stream was changed to a wide fog stream for 30 seconds. The procedure was repeated an additional time.
